### 2016-12-08 manuscript/manuscript-speed-accuracy.tex
Added licence, tidied README...

In the following study, we explore factors that may be indicative of software accuracy. This, in our opinion, should be one of the prime reasons for selecting a software tool. We have mined the large and freely accessible PubMed database \cite{Sayers2010-vm} for benchmarks of computational biology software, and manually extracted accuracy and speed rankings for \textbf{{\color{red}243}} software packages. For each software tool, we have collected measures that may be predictive of accuracy, and may be subjectively employed by the researcher community as a proxy for software quality. These include relative speed, relative age, the productivity and impact of the corresponding authors, journal impact and the number of citations.

\section*{Results}
We have collected relative accuracy and speed ranks for \textbf{{\color{red}243}} distinct software tools. This software has been developed for solving a broad cross-section computational biology tasks. These include software for homology search \cite{Freyhult2007-et}, genome sequence analysis (e.g. read mapping or sequence assembly) \cite{Junemann2014-mb,Tran2014-pe,Zhang2011-nd,Abbas2014-gu,Bao2011-lv,Caboche2014-lj,Kleftogiannis2013-wi,Hatem2013-cs,Schbath2012-ob,Ruffalo2011-rl,Yang2013-aj,Holtgrewe2011-fd,Rackham2015-ag,Huang2015-wu}, multiple sequence alignment \cite{Thompson2011-rf,Nuin2006-nk,Pais2014-sr,Pervez2014-zp,Liu2010-rp}, cell tracking \cite{Maska2014-ak}, transcriptome analysis \cite{Li2012-wr,Lu2013-fs,Liu2014-kz,Kumar2016-xz}, RNA interaction prediction \cite{Pain2015-gr}, protein interactions \cite{Tikk2010-qd}, protein structure prediction \cite{Kolodny2005-ry,Wallner2005-qi}, epistasis \cite{Shang2011-vy},  metagenomic analysis \cite{Lindgreen2016-tt,Bazinet2012-wf}, repetitive sequence prediction \cite{Saha2008-kd}, proteomics \cite{Lange2008-pt,Yang2009-oc} and phylogenetics \cite{Liu2011-pz,Yang2011-dv,Oscamou2008-md,Bayzid2013-hc,Liu2009-lx}. Each software tool was benchmarked in at least one of \textbf{{\color{red}43}} publications that satisfy the Boulesteix criteria \cite{Boulesteix2013-vb}. In brief, the Boulesteix criteria are: 1. the main focus of the article is a benchmark. 2. the authors are reasonably neutral. 3. the test data and evaluation criteria are sensible.

@@

We found that author reputation metrics, journal impacts and the age of methods were \textbf{not} significantly correlated with either method accuracy or speed (see Figure~\ref{fig:allfactors}). The strongest association was between accuracy and journal impact factor (Spearman’s rho = 0.1, P-value = 0.16). A linear model of these parameters and accuracy also failed to identify a correlation between these (accuracy: R2 = -0.03, P-value = 0.94; speed: R2 = 0.04, P-value = 0.66).

\begin{equation*}
\begin{split}
speed=& c_9+c_{10}\times accuracy+c_{11}\times JIF+c_{12}\times H5+\\
& c_{13}\times citations+c_{14}\times Hindex+\\
& c_{15}\times Mindex+c_{16}\times relativeAge+\\
& c_{17}\times relativeCitations
\end{split}
\end{equation*}
To further validate this result, we compute a correlation between speed and accuracy for each benchmark and used weighted sum Z-tests \cite{Zaykin2011-tj}. This also failed to identify a significant relationship (sum Z=-0.1, P-value=0.5).

@@

@@

\section*{Discussion}
We have gathered data on the relative speed and accuracies of \textbf{{\color{red}243}} bioinformatic methods from \textbf{{\color{red}43}} benchmarks that were published between 2005 and 2016. We show that there is an \textbf{under-representation of software that has both intermediate levels of accuracy and speed}. There may be a number of factors that drive this phenomena. One likely explanation is that \textbf{bioinformatic software tools suffer from a form of publication bias} \cite{Boulesteix2015-am}. Our community of developers, reviewers and editors may be unwilling to publish software that is neither the fastest nor the most accurate (we have anecdotal evidence to support this, N=1). If correct, this is unfortunate. Some problems that lack fast and accurate solutions force researchers to make unnecessary compromises in terms of accuracy and time. If our hypothesis that this underrepresentation is due to publication bias is valid then why is there an enrichment of slow and inaccurate software (P-value=0.05, empirical distributions from permutation tests)? How could these methods be published? We have found that slow and inaccurate software is generally published earlier than alternative methods (P=0.007, one-tailed Wilcoxon test), therefore comparisons were not required to publish these tools and the accuracies may have been over-optimistically reported \cite{Boulesteix2015-nc}.

@@

We propose that the full spectrum of software tool accuracies and speeds serve a useful purpose to the research community. Like negative results, if honestly reported, illustrate to the research community that certain approaches are not practical research avenues \cite{Ioannidis2005-xh,Workman1999-au,Rivas2000-fb}. The current practises of publishers, editors, reviewers and authors of software tools appear to have  deprived our community of tools for building effective and productive workflows.

@@

A score is computed for each candidate article by summing the log-odds scores for the words in title and abstract, i.e. $s(a)=\sum_i^Nlo(w_i)$. The high scoring candidate articles are then manually evaluated against the ALB-criteria. Accuracy and speed ranks are extracted from the articles that meet the criteria, and these are also added to the set of training articles. The evaluated candidate articles that do not meet the ALB-criteria are incorporated into the set of background articles. This process is iterated a number of times and has resulted in the identification of \textbf{{\color{red}43}} benchmark articles, that contain \textbf{{\color{red}102}} different benchmarks, together these rank \textbf{{\color{red}243}} distinct software packages.

@@

\textbf{Data extraction:} for each article that met the ALB-criteria and contained data on both the accuracy and speed from their tests we extracted ranks for each method. Many articles contained multiple benchmarks, in these cases we selected a range of these, the provenance of which is stored with the accuracy metric and raw speed and accuracy rank data for each method. In line with rank-based statistics, the cases where methods were tied are resolved by using a midpoint rank (e.g. if method 3 and 4 are tied, the rank 3.5 is used) \cite{Mann1947-re}. Each rank extraction was independently verified by at least one other co-author to ensure both the provenance of the data could be established and that the ranks were correct. The ranks for each benchmark were then normalised to lie between 0 and 1 using the formula $\frac{r-1}{n-1}$ where ‘r’ is a method’s rank and ‘n’ is the number of methods in the benchmark. For methods that were benchmarked multiple times with multiple metrics (e.g. BWA is evaluated in 6 different articles \cite{Bao2011-lv,Caboche2014-lj,Hatem2013-cs,Schbath2012-ob,Ruffalo2011-rl,Holtgrewe2011-fd}) a mean normalised rank is used to summarise the performance.

@@

@@

\begin{figure*}
\includegraphics[scale=0.7]{figure1.pdf}
\caption{\textbf{A.} A heatmap indicating the relationships between proposed predictors of software quality. Spearman’s rho is used to infer correlations between metrics such as the H and M indices of corresponding authors, number of citations, journal impact factors and H5 indices, the year and relative age of software and the mean relative rankings of software speed and accuracy. Red colours indicate a high positive correlation, blue colours indicate a high negative correlation. Correlations with a P-value less than 0.05 are indicated with a ‘X’. The dashed rectangular area is illustrated in more detail in \textbf{B}, the bold square is shown in more detail in \textbf{Figure 2}.The dendrogram was computed using the default ‘heatmap.2 function in R, which computes a Euclidean distance matrix and a complete-linkage hierarchical clustering. \textbf{B.} A barplot illustrating the correlation as measured by Spearman’s rho between and normalised accuracy ranks and potential factors that may be predictive of accuracy. In order to give an appreciation of the difference between the observed effect-sizes and significant effect sizes, we generated 10,000 permuted accuracy ranks for each benchmark and recorded Spearman’s rho for the significant correlations (P<0.05). These values are marked with “x”s in the barplot.}
\label{fig:allfactors}
\end{figure*}

\begin{figure*}
\includegraphics[scale=0.55]{figure2.pdf}
\caption{A heatmap indicating the relative paucity or abundance of software in the range of possible accuracy and speed rankings. Red colours indicate an abundance of software tools in an accuracy and speed category, while blue colours indicate scarcity of software in an accuracy and speed category. The abundance is quantified using a Z-score computation for each bin, this is derived from 10,000 random permutations of the speed and accuracy ranks from each benchmark. Mean normalised ranks of accuracy and speed have been binned into 100 classes (a $10\times10$ grid) that range from comparatively slow/inaccurate to comparatively fast/accurate. Z-scores with a P-value less than 0.05 are indicated with a ‘X’.}
\label{fig:speedaccuracy}
\end{figure*}

>>>>

In the following study, we explore factors that may be indicative of software accuracy. This, in our opinion, should be one of the prime reasons for selecting a software tool. We have mined the large and freely accessible PubMed database \cite{Sayers2010-vm} for benchmarks of computational biology software, and manually extracted accuracy and speed rankings for \textbf{{\color{black}243}} software packages. For each software tool, we have collected measures that may be predictive of accuracy, and may be subjectively employed by the researcher community as a proxy for software quality. These include relative speed, relative age, the productivity and impact of the corresponding authors, journal impact and the number of citations.

\section*{Results}
We have collected relative accuracy and speed ranks for \textbf{{\color{black}243}} distinct software tools. This software has been developed for solving a broad cross-section computational biology tasks. These include software for homology search \cite{Freyhult2007-et}, genome sequence analysis (e.g. read mapping or sequence assembly) \cite{Junemann2014-mb,Tran2014-pe,Zhang2011-nd,Abbas2014-gu,Bao2011-lv,Caboche2014-lj,Kleftogiannis2013-wi,Hatem2013-cs,Schbath2012-ob,Ruffalo2011-rl,Yang2013-aj,Holtgrewe2011-fd,Rackham2015-ag,Huang2015-wu}, multiple sequence alignment \cite{Thompson2011-rf,Nuin2006-nk,Pais2014-sr,Pervez2014-zp,Liu2010-rp}, cell tracking \cite{Maska2014-ak}, transcriptome analysis \cite{Li2012-wr,Lu2013-fs,Liu2014-kz,Kumar2016-xz}, RNA interaction prediction \cite{Pain2015-gr}, protein interactions \cite{Tikk2010-qd}, protein structure prediction \cite{Kolodny2005-ry,Wallner2005-qi}, epistasis \cite{Shang2011-vy},  metagenomic analysis \cite{Lindgreen2016-tt,Bazinet2012-wf}, repetitive sequence prediction \cite{Saha2008-kd}, proteomics \cite{Lange2008-pt,Yang2009-oc} and phylogenetics \cite{Liu2011-pz,Yang2011-dv,Oscamou2008-md,Bayzid2013-hc,Liu2009-lx}. Each software tool was benchmarked in at least one of \textbf{{\color{black}43}} publications that satisfy the Boulesteix criteria \cite{Boulesteix2013-vb}. In brief, the Boulesteix criteria are: 1. the main focus of the article is a benchmark. 2. the authors are reasonably neutral. 3. the test data and evaluation criteria are sensible.

@@

\begin{figure*}[h]
\includegraphics[width=\textwidth]{figure1.pdf}
\caption{\textbf{A.} A heatmap indicating the relationships between proposed predictors of software quality. Spearman’s rho is used to infer correlations between metrics such as the H and M indices of corresponding authors, number of citations, journal impact factors and H5 indices, the year and relative age of software and the mean relative rankings of software speed and accuracy. Red colours indicate a high positive correlation, blue colours indicate a high negative correlation. Correlations with a P-value less than 0.05 are indicated with a ‘X’. The dashed rectangular area is illustrated in more detail in \textbf{B}, the bold square is shown in more detail in \textbf{Figure 2}.The dendrogram was computed using the default ‘heatmap.2 function in R, which computes a Euclidean distance matrix and a complete-linkage hierarchical clustering. \textbf{B.} A barplot illustrating the correlation as measured by Spearman’s rho between and normalised accuracy ranks and potential factors that may be predictive of accuracy. In order to give an appreciation of the difference between the observed effect-sizes and significant effect sizes, we generated 10,000 permuted accuracy ranks for each benchmark and recorded Spearman’s rho for the significant correlations (P<0.05). These values are marked with “x”s in the barplot.}
\label{fig:allfactors}
\end{figure*}

We found that author reputation metrics, journal impacts and the age of methods were \textbf{not} significantly correlated with either method accuracy or speed (see Figure~\ref{fig:allfactors}). The strongest association was between accuracy and journal impact factor (Spearman’s rho = 0.1, P-value = 0.16). A linear model of these parameters and accuracy also failed to identify a correlation between these (accuracy: R2 = -0.03, P-value = 0.94; speed: R2 = 0.04, P-value = 0.66). To further validate this result, we compute a correlation between speed and accuracy for each benchmark and used weighted sum Z-tests \cite{Zaykin2011-tj}. This also failed to identify a significant relationship (sum Z=-0.1, P-value=0.5).

@@

\begin{figure*}
%\includegraphics[scale=0.55]{figure2.pdf}
\includegraphics[width=\textwidth]{figure2.pdf}
\caption{A heatmap indicating the relative paucity or abundance of software in the range of possible accuracy and speed rankings. Red colours indicate an abundance of software tools in an accuracy and speed category, while blue colours indicate scarcity of software in an accuracy and speed category. The abundance is quantified using a Z-score computation for each bin, this is derived from 10,000 random permutations of the speed and accuracy ranks from each benchmark. Mean normalised ranks of accuracy and speed have been binned into 100 classes (a $10\times10$ grid) that range from comparatively slow/inaccurate to comparatively fast/accurate. Z-scores with a P-value less than 0.05 are indicated with a ‘X’.}
\label{fig:speedaccuracy}
\end{figure*}

@@

\section*{Discussion}

We have gathered data on the relative speed and accuracies of \textbf{{\color{black}243}} bioinformatic methods from \textbf{{\color{black}43}} benchmarks that were published between 2005 and 2016. We show that there is an under-representation of software that has both intermediate levels of accuracy and speed. There may be a number of factors that drive this phenomena. One likely explanation is that bioinformatic software tools suffer from a form of publication bias \cite{Boulesteix2015-am}. Our community of developers, reviewers and editors may be unwilling to publish software that is neither the fastest nor the most accurate (we have anecdotal evidence to support this, N=1). If correct, this is unfortunate. Some problems that lack fast and accurate solutions force researchers to make unnecessary compromises in terms of accuracy and time. If our hypothesis that this underrepresentation is due to publication bias is valid then why is there an enrichment of slow and inaccurate software (P-value=0.05, empirical distributions from permutation tests)? How could these methods be published? We have found that slow and inaccurate software is generally published earlier than alternative methods (P=0.007, one-tailed Wilcoxon test), therefore comparisons were not required to publish these tools and the accuracies may have been over-optimistically reported \cite{Boulesteix2015-nc}.

@@

We propose that the full spectrum of software tool accuracies and speeds serve a useful purpose to the research community. Like negative results, if honestly reported, illustrate to the research community that certain approaches are not practical research avenues \cite{Ioannidis2005-xh,Workman1999-au,Rivas2000-fb}. The classes of tools that we find in the under-represented accuracy and speed region include some of the most used tools in bioinformatics, such as the homology search tools NCBI-BLAST \cite{Altschul1990-ht,Altschul1997-ga} and HMMER \cite{Eddy:2009} and genome assembly and mapping tools such as AbySS \cite{Simpson:2009}, MAQ \cite{Li:2008} and Newbler \cite{Margulies:2005}. The current practises of publishers, editors, reviewers and authors of software tools therefore deprive our community of tools for building effective and productive workflows.

@@

A score is computed for each candidate article by summing the log-odds scores for the words in title and abstract, i.e. $s(a)=\sum_i^Nlo(w_i)$. The high scoring candidate articles are then manually evaluated against the ALB-criteria. Accuracy and speed ranks are extracted from the articles that meet the criteria, and these are also added to the set of training articles. The evaluated candidate articles that do not meet the ALB-criteria are incorporated into the set of background articles. This process is iterated a number of times and has resulted in the identification of \textbf{{\color{black}43}} benchmark articles, that contain \textbf{{\color{black}102}} different benchmarks, together these rank \textbf{{\color{black}243}} distinct software packages.

@@

\textbf{Data extraction and processing:} for each article that met the ALB-criteria and contained data on both the accuracy and speed from their tests we extracted ranks for each method. Many articles contained multiple benchmarks, in these cases we selected a range of these, the provenance of which is stored with the accuracy metric and raw speed and accuracy rank data for each method. In line with rank-based statistics, the cases where methods were tied are resolved by using a midpoint rank (e.g. if method 3 and 4 are tied, the rank 3.5 is used) \cite{Mann1947-re}. Each rank extraction was independently verified by at least one other co-author to ensure both the provenance of the data could be established and that the ranks were correct. The ranks for each benchmark were then normalised to lie between 0 and 1 using the formula $\frac{r-1}{n-1}$ where ‘r’ is a method’s rank and ‘n’ is the number of methods in the benchmark. For methods that were benchmarked multiple times with multiple metrics (e.g. BWA is evaluated in 6 different articles \cite{Bao2011-lv,Caboche2014-lj,Hatem2013-cs,Schbath2012-ob,Ruffalo2011-rl,Holtgrewe2011-fd}) a mean normalised rank is used to summarise the performance.

@@

The linear models that we used to test for relationships between speed, accuracy and the above measures are:

\begin{equation*}
\begin{split}
speed=& c_0+c_{1}\times accuracy+c_{2}\times JIF+c_{3}\times H5+\\
& c_{4}\times citations+c_{5}\times Hindex+\\
& c_{6}\times Mindex+c_{7}\times relativeAge+\\
& c_{8}\times relativeCitations
\end{split}
\end{equation*}

@@
